Seeq: a library for inexact DNA sequence matching

### 2015-03-31 doc/document.tex
Enabled inverse mode again.
First edits on the text.

\section{Motivation:}
Searching specific sequences over DNA reads is an elemental procedure
of biological data analysis. Experimental DNA samples have usually
been engineered to contain pattern sequences at the positions where
operations such as adaptor removal, sequence trimming or barcode
extraction have to be performed. The experimental data contain errors
that must be tolerated in order to identify these patterns at their
correct positions. This process is straightforward in data produced by
second generation sequencing machines, in which the error rates lie
below 2\%. Short sequences with small error tolerance suffice to
identify the patterns accurately in these circumstances. However,
third generation sequencing technologies promise longer reads with
error rates around 15\%. Consequently, this opens a new scenario in
which both the pattern length and the relative error tolerance will
have to be dramatically increased.
\section{Results:}
In this paper we present seeq, an inexact pattern matching algorithm
optimized for DNA sequence matching with almost constant running
time. We benchmark seeq against the most widely used algorithms for
inexact pattern matching. The results show that seeq scales better and
becomes significantly faster as the error rate increases.
\section{Availability:}
Seeq is available in three forms: Linux software, C Library and Python
module. The source code and the C/Python libraries are available for
download at
\href{http://github.com/ezorita/seeq}{http://github.com/ezorita/seeq}.
\section{Contact:} \href{guillaume.filion@gmail.com}{guillaume.filion@gmail.com}
\end{abstract}

@@

Inexact string matching is an old problem that dates back to the
sixties with the definition of the edit distance \citep{Lev65}. The
first algorithms used to solve this problem were based on dynamic
programming \citep{NW70}. Although the algorithms were not very
efficient, they are still widely used to align biological sequences
due to their simplicity and flexibility. Later on, more sophisticated
algorithms derived from the Boyer-Moore search method \citep{BM77}
were designed to search patterns on long strings allowing
differences. Such algorithms are much faster than the dynamic
programming approach because they do not compute all the relative
distances. Instead, the pattern is pre-processed to design a search
strategy which skips parts of the text that do not contain any
match. The jumping strategy is very efficient for perfect matches or
small edit distances because the non-matching condition is inferred
with only a few comparisons. In this scenario, the number of
operations per input character is ridiculously small given that large
areas of the text are never processed. However, other scenarios are
not so favorable for this technique. For instance, if the proposed
pattern matches in almost any text position, the performance degrades
severely since the algorithm will not be able to jump so
often. Besides, for high mismatch ratios the search strategy becomes
complex and requires more comparisons to determine a non-matching
condition, specially when the alphabet is as small as in the task of
matching DNA sequences.

Sequencing data produced by the Illumina platform \citep{Mar05}
typically has 1-2\% error rate mainly consisting of substitutions
\citep{Doh08}. Therefore one expects to find all the matching strings
within a small edit distance relative to the original pattern
length. Moreover, if such patterns are produced artificially
(e.g. sequencing adaptors or barcodes), it is reliable to use short 
sequences that are not present in the target genome. On the other
side, the PacBio platform produces reads two orders of magnitude
longer than Illumina, but at the expense of a much higher error rate
(15\%) consisting principally of insertions and deletions
\citep{Eid09}. This high error rate dramatically increases the false
positive rate unless the pattern sequence is extended
accordingly. This brings us to a new scenario with unusually long
patterns and edit distances that lie on the same order of magnitude.

In this article we present \emph{seeq}, an algorithm based on a
deterministic finite automaton (DFA) structure. The automaton is built
as the input text is processed and has been optimized for DNA sequence
matching. We benchmark seeq against other widely used inexact string
matching software. The results show that seeq scales better and
becomes significantly faster as the error rate increases.

@@

The algorithm proposed is the DFA representation of a dynamic
programming algorithm. The goal is to find the occurrences of a
pattern $P$ in a text $T$ that are $k$-differences or less. The
lengths of $P$ and $T$ are $m$ and $n$, respectively. The score matrix
$\mathbf{C} \in \mathbb{N}^{m\times n}$ is initialized so as that any
text position is a potential alignment start, i.e. $\mathbf{C}(0,j)=0$
and $\mathbf{C}(i,0) = i$. After the initialization, the score matrix
is computed and filled column-wise. Note that the standard dynamic
programming update process can be represented as
$\mathbf{c}_j=f(\mathbf{c}_{j-1},T_j)$. In effect, the value of the 
$j$-th column $\mathbf{c}_{j}$ is a deterministic transition that only
depends on $\mathbf{c}_{j-1}$ and the character $T_j$. We take
advantage of this property to build a deterministic automata in which
the different states represent the set of non-equivalent alignment
columns and the transitions between states are dictated by the input
characters.

To make the alignment algorithm efficient, only the significant
($\mathbf{C}(i,j) \le k$) parts of the column are computed. The DFA
states $S$ have a vector representation
$\mathbf{s} = \mbox{dif}(\mathbf{c}_j)$ generated by differentially
encoding the $j$-th column values as 
$\mathbf{s}(i)=\mathbf{C}(i+1,j)-\mathbf{C}(i,j)+1$ for
$i=0\ldots m-1$. This is done for two reasons: (i) the alphabet of
$\mathbf{s}$ is small \{0,1,2\}, therefore the alignment columns can be
efficiently stored in a ternary trie, and (ii) the inverse process to
recompute the original column ($\mathbf{c}_j=\mbox{undif}(\mathbf{s})$)
is performed without further information, provided that $\mathbf{C}(0,j)=0$.

@@

The DFA construction process is illustrated in Figure. X. The
dynamic programming algorithm starts at the DFA state
$S = S_0$. For any input character $T_j$, being $S$
the active DFA state, the algorithm recursively proceeds as follows:
\begin{itemize}
\item If $S$ has an outgoing path (to $S_q$) with
label $T_j$, then $S = S_q$.
\item Otherwise, the vector $\mathbf{s}_j$ is computed as
$\mathbf{s}_j=\mbox{dif}(f(\mbox{undif}(\mathbf{s}), T_j))$.
If a state represented by $\mathbf{s}_j$ already exists ($S_q$), $S$
is connected to such state with a path labeled $T_j$ and $S = S_q$. On
the other hand, if $\mathbf{s}_j$ does not exist, a new DFA state
$S_j$ is created and both nodes are connected in the same
way, then $S = S_j$.
\end{itemize}
After several steps, the DFA graph is densely connected and the
complexity of the algorithm boils down to direct DFA state
transitions.

@@

We have benchmarked seeq against two competitive inexact string
matching algorithms: agrep \citep{Wu92} and nr-grep \citep{Nav01}. We
used two real sequencing datasets for benchmarking. Both datasets contain
10 million reads but differ in read length. Dataset 1 consists random
Drosophila genome reads of length 100. Dataset 2 reads are 50
nucleotides long, from which the first 30 nucleotides are constant and
the last 20 are random (barcodes). The software were set to match
random patterns ranging from 5 to 50 nucleotides with a random error
tolerance up to 40\%. We would like to point out that agrep limits the
number of differences to 8, therefore the comparison shown here is
only fair between seeq and nr-grep.

Figure \ref{fig:results} shows the running time of each execution as
a function of the relative error tolerace ($k/m$). For low error
rates, both agrep and nr-grep perform better than seeq, with a maximum
5-fold advantage when $k=1$ and $m > 25$. On the other hand, seeq is
up to two orders of magnitude faster than nr-grep and one order of
magnitude faster than agrep for $k/m > 0.15$. The similarity between
the two datasets indicates that neither software is particularly
favored when the processed data has a repetitive structure. The worst
to best case ratio also speaks strongly in favor of seeq (5), which
achieves almost constant running time. This value is one and almost
three orders of magnitude higher for agrep (50) and nr-grep (1000),
respectively.

\begin{figure}[!tpb]
\centerline{\includegraphics[scale=0.5]{results.pdf}}
\caption{Benchmark between seeq, agrep and nr-grep.}
\label{fig:results}
\end{figure}

@@

In this article we have introduced and benchmarked seeq, an inexact
DNA sequence matching algorithm. We have measured and compared the
running time of seeq and two competitive software in different
configurations. The results conclude that seeq is faster in higher
error regions with a worst case speedup of almost 100. We attribute
the higher efficiency of seeq to the fact that it has been
specifically designed to match DNA sequences. Thus, we could take
advantage of algorithms that are not feasible in general string
matchers because they only scale well when the size of the alphabet is
reduced.

>>>>

\section{Motivation:}
Searching specific sequences in DNA is one of the most common tasks
in the analysis of biological data. Experimental samples often contain
known sequences at the positions where adaptor removal, sequence
trimming or barcode extraction have to be performed. Experimental data
contain errors that must be tolerated in order to identify these
patterns. For reads produced by second generation sequencing,
where the error rates lie below 2\%, short sequences with small error
tolerance suffice to identify the patterns. However, third generation
sequencing technologies promise longer reads at the cost of 10-fold
higher error rates. In this scenario, both the pattern length and the
relative error tolerance have to be dramatically increased.
\section{Results:}
Here we present seeq, an algorithm for pattern matching with errors.
Seeq is optimized for DNA sequences and shows very little increase in
running time as error tolerance increases. We benchmark seeq against
the most widely used algorithms for inexact pattern matching. The
results show that seeq scales better and becomes significantly faster
as the error rate increases.
\section{Availability:}
Seeq is available in three forms: a Linux software, a C library and a 
Python module. The source code and the C/Python libraries are available
for download at
\href{http://github.com/ezorita/seeq}{http://github.com/ezorita/seeq}.
\section{Contact:} \href{guillaume.filion@gmail.com}{guillaume.filion@gmail.com}
\end{abstract}

@@

Inexact string matching dates back to the origin of computer sciences
\citep{Lev65}. The first algorithms addressing this problem were based
on dynamic programming \citep{NW70}. They are still widely used to
align biological sequences due to their simplicity and flexibility,
More sophisticated algorithms derived from the Boyer-Moore search
method \citep{BM77} were later designed to search patterns on long
strings allowing errors. Such algorithms are faster than the dynamic
programming approach because they do not compute all the relative
distances. Instead, the pattern is pre-processed to design a search
strategy which skips parts of the text that do not contain any
match. The jumping strategy is very efficient for perfect matches or
small edit distances because non-matching conditions can be inferred
with only a few comparisons. In those conditions, the number of
operations per input character is small because large areas of the
text are never processed. However, when the error rate is high, the
search strategy requires more comparisons to identify a non-matching
condition. The drop in performance is even more accute when the alphabet
is as small as in the case of DNA sequences.

Illumina reads typically have a 1-2\% error rate \citep{Mar05},
mainly consisting of substitutions \citep{Doh08}, which are ideal
conditions for the Boyer-Moore algorithm. However, third generation
sequencing technologies such as the PacBio platform now come of age.
They produce reads two orders of magnitude longer than Illumina, at
the expense of a 10-fold higher error rate (15\%) consisting mostly
of insertions and deletions \citep{Eid09}.  
This high error rate dramatically increases the false
positive rate unless the pattern sequence is extended
accordingly. This raises a new scenario with long patterns and
large edit distances.

Here we present \emph{seeq}, an algorithm using a dynamic version of
deterministic finite automata (DFA). Instead of using standard
construction procedures, the dynamic DFA of seeq is built as the input
text is processed. We benchmark seeq against other widely used inexact
string matching software. The results show that seeq scales better and
becomes significantly faster as the error rate increases.

@@

We use a DFA representation of a dynamic programming algorithm.
We search the occurrences of the pattern $P$ of length $m$
in a text $T$ of length $n$, with up to $k$ differences.
The edit matrix $\mathbf{C} \in \mathbb{N}^{m\times n}$ is initialized
with $\mathbf{C}(0,j)=0$ and $\mathbf{C}(i,0) = i$, and then updated
column-wise with the Needleman-Wunsch recusrion. Note that this process
can be represented as $\mathbf{c}_j=f(\mathbf{c}_{j-1},T_j)$,
\textit{i.e.} the value of the $j$-th column $\mathbf{c}_{j}$ is a
deterministic transition that only depends on $\mathbf{c}_{j-1}$ and the
character $T_j$. We take advantage of this property to build a
deterministic automaton where the states are the columns of the matrix
and the transitions are dictated by the input characters.
We represent the states of the DFA by the differential encoding
of the $j$-th column, $\mathbf{s}(i)=\mathbf{C}(i+1,j)-\mathbf{C}(i,j)+1$
so that the state can be stored in a ternary trie ($\mathbf{s}(i)$
can take only 3 values).

@@

The DFA construction process is illustrated in Figure. X. The
dynamic programming algorithm starts at the DFA state
$S = S_0$. For any input character $T_j$, being $S$
the active DFA state, the algorithm recursively proceeds as follows:
\begin{enumerate}
\item If $S$ has an outgoing path (to $S_q$) with
label $T_j$, then set $S = S_q$.
\item Otherwise, compute the column of the edit matrix corresponding
to $f(S, T_j)$. Case 2a: if this state already exists ($S_q$), connect
$S$ to it with a path labeled $T_j$ and set $S = S_q$. Case 2b: if
the state does not exist, create it, and do as in in the previous
case 2a.
\end{enumerate}

After several steps, the graph of the DFA is densely connected and
the steps of state and edge creation are almost never performed. 
The complexity of the algorithm is then reduced to direct DFA state
transitions.

@@

We have benchmarked seeq against two standard inexact string
matching algorithms: agrep \citep{Wu92} and nr-grep \citep{Nav01}. We
used two sequencing datasets for benchmarking. Both datasets contain
10 million reads but differ in read length. Dataset 1 consists of random
\textit{Drosophila} genome reads of length 100. The reads of dataset 2
are 50 nucleotides long, from which the first 30 nucleotides are constant
and the last 20 are random (barcodes). The software were set to match
random patterns ranging from 5 to 50 nucleotides with a random error
tolerance up to 40\%. Importantly, agrep limits the number of differences
to 8, which is a serious limitation and makes the comparison fair only between seeq and nr-grep.

The running time of each execution is plotted against the relative
error tolerace ($k/m$) on figure \ref{fig:results}. For low error
rates, both agrep and nr-grep perform better than seeq, with a maximum
5-fold advantage when $k=1$ and $m > 25$. On the other hand, seeq is
up to two orders of magnitude faster than nr-grep and one order of
magnitude faster than agrep for $k/m > 0.15$. The similarity between
the two datasets indicates that neither software is favored when the
data has a repetitive structure. Also note that seeq 
achieves near-constant running time.

\begin{figure}[!tpb]
\centerline{\includegraphics[scale=0.5]{results.pdf}}
\caption{Benchmark between seeq, agrep and nr-grep. Each dot represents
the running time of one search with a random pattern and a random
edit distance. Note the logarithmic scale on the y axis.}
\label{fig:results}
\end{figure}

@@

The benchmark shows that seeq is faster in higher error regions with
an excellent worst case. Actually, the running time of seeq depends
very little on the error tolerance. The higher efficiency of seeq
is attributed to the underlying dynamic DFA structure. Here, the key
insight was to take advantage of the limited size of the DNA alphabet.
This strategy is usually not efficient in general string matching
problems.

In practice, most pattern searches in DNA reads are performed from
scripts and pipelines. To ease the use of seeq in existing software,
we have ensured that it is comliant with Linux pipes. Also, we have
created a Python module that can be imported and integrated seamlessly,
as it can be called on Python string objects that can be further
manipulated. We also provide a API in C for the cases that optimal
performance is required.
